---
layout: topic
title: Introducing R
author: Data Carpentry contributors
minutes: 45
---


```{r, echo=FALSE, purl=FALSE, message = FALSE}
source("setup_wehi.R")
```

------------

> ## Learning Objectives
>
> * Begin to appreciate the power of using R 
> * Understand some of the backbones of R (functions, objects, vectors, matrices)

------------

## The Awesome World of R

_Start by showing an example of a script that generates a beautiful graph_

```{r, purl=FALSE}
pairs(iris[1:4], main = "Anderson's Iris Data -- 3 species", pch = 21, bg = c("red", "green3", "blue")[unclass(iris$Species)])
```

* Point to the different parts:
  - a function
  - the assignment operator `<-`
  - the `=` for arguments
  - the comments `#` and how they are used to document function and its content
  - the `$` operator
* Point to indentation and consistency in spacing to improve clarity

Two popular style guides for R are [Hadley Wickham's](http://adv-r.had.co.nz/Style.html) and
[Google's](https://google-styleguide.googlecode.com/svn/trunk/Rguide.xml).


## Objects in R

```{r, echo=FALSE, purl=TRUE}
### Objects in R

```

R can be used as a calculator:

```{r, purl=FALSE}
3 + 5
12/7
sqrt(16)
```

R can assign _values_ to _objects_ that will be stored:

```{r, purl=FALSE}
weight_kg <- 55   # Note: `weight_kg` is different from `Weight_kg`
```

Display values of an object using () or re-type object name:

```{r, purl=FALSE}
(weight_kg <- 55)
weight_kg
```

Perform calculations on stored objects, e.g. convert weight_kg to pounds (i.e. multiply by 2.2):

```{r, purl=FALSE}
2.2 * weight_kg
```

A variable's value can be replaced by a new one:

```{r, purl=FALSE}
weight_kg <- 57.5
```

Create a new variable based on another stored variable:

```{r, purl=FALSE}
weight_lb <- 2.2 * weight_kg
```

## Vectors

```{r, echo=FALSE, purl=TRUE}
### Vectors and data types

```

A vector is a list of values for one variable. Vector values must be the same type e.g. all numbers. 

Create variable 'x' with numbers between 2 and 4:

```{r, purl=FALSE}
x <- 2:4         
```

Create variable 'y' with numbers 3, 6, 8 (using 'c' to 'concatenate'/'combine'):

```{r, purl=FALSE}
y <- c(3,6,8)     
```

Perform a function on a stored variable:

```{r, purl=FALSE}
2 * y     
x + y     # add two vectors together
```

Compare vectors:

```{r, purl=FALSE}
x < y
x == c(2, 5, 10)
```

A vector can contain characters:

```{r, purl=FALSE}
celltype <- c("eosinophils", "basophils", "neutrophils")
```

Extract individual values from vectors using [position]:

```{r, purl=FALSE}
celltype[3]
```

Replace values in an existing vector:

```{r, purl=FALSE}
celltype[2] <- "dc"
```

Add more values to the end of a vector:

```{r, purl=FALSE}
celltype <- c(celltype,"monocytes")
```

Retrieve multiple values from vector:

```{r, purl=FALSE}
celltype[3:4]
celltype[c(1,3,5)]
```

Assign names to vector values, then use names to retrieve these values:

```{r, purl=FALSE}
y <- c(3,6,8) 
names(y) <- c("DCs", "monocytes", "neutrophils")
y["neutrophils"]
```


R has many functions that allow you to inspect the content of a vector:

```{r, purl=FALSE}
length(celltype)   # how many elements are there?
class(celltype)    # what is the class (type of element)?
```

The function `str()` gives an overview of the object and the elements it
contains:

```{r, purl=FALSE}
str(celltype)
```

## Matrices

```{r, echo=FALSE, purl=TRUE}
### Matrices

```

A matrix is a collection of values arranged in a two-dimensional layout (row x column) 

Create a matrix containing numbers 1:15 with 5 rows x 3 columns:

```{r, purl=FALSE}
x <- matrix(data = 1:15, nrow = 5, ncol = 3)
x
```

Access values from matrix using [row,column]:

```{r, purl=FALSE}
x[2,3]    # row 2, column 3
x[2,]     # all of row 2
x[,3]     # all of column 3
```

Vectors and matrices are some of the many **data structures** that R uses. 

Other important ones are lists (`list`), matrices, data frames (`data.frame`) and factors (`factor`).

We are now going to use our "cytokines" dataset to explore the `data.frame` and 'factor' data structures.
