---
layout: topic
title: Data Analysis
author: Data Carpentry contributors
minutes: 60
---

```{r, echo=FALSE, purl=FALSE, message = FALSE}
source("setup_wehi.R")
```

------------

> ## Learning Objectives
>
> * Build an understanding of the concepts of a `data.frame`
> * Learn to load external data (CSV files) in memory using
>  (`cytokines.csv`) as an example
> * Explore structure and the content of the data
> * Understand what are factors and how to manipulate them
> * Introduce exploratory plots, statistical analyses

------------
# Cytokine Data

```{r, echo=FALSE, purl=TRUE}
### Cytokine Data
```

Serum cytokine levels were measured in "control" and "infected" patients.

The dataset is stored as a `csv` file: each row holds information for a single patient, and the columns represent:

| Column           | Description                        |
|------------------|------------------------------------|
| patient_id       | unique id for each patient         |
| group            | status of the patient              |
| IL17F            | concentration of IL17F             |
| IL17A            | concentration of IL17A             |
| GMCSF            | concentration of GMCSF             |
| IL2              | concentration of IL2               |
| CXCL9            | concentration of CXCL9             |
| CXCL10           | concentration of CXCL10            |
| ENA78            | concentration of ENA78             |
| IL4              | concentration of IL4               |


### Create Folders and Set Working Directory

```{r, echo=FALSE, purl=TRUE}
### Create Folders and Set Working Directory
```

- Create a new folder in documents, called "r_scripts"
- Locate "r_scripts" in RStudio ('Files', lower right window)
- Set as working directory `setwd("~/Documents/r_scripts")` or use 'More' icon
- Create new folder `dir.create("data")` or use 'New Folder' icon

### Load Cytokine Data
```{r, echo=FALSE, purl=TRUE}
### Load Cytokine Data
```

Download cytokines data:

```{r, eval=FALSE, purl=TRUE}
download.file("http://files.figshare.com/2443264/cytokines_3groups.csv",
              "data/cytokines_3groups.csv")
```

Load the data:

```{r, eval=TRUE,  purl=FALSE}
cytokines <- read.csv('data/cytokines_3groups.csv')
```

Type variable `cytokines` to view, or wrap assignment in parentheses to assign and display at the same time.

```{r, eval = TRUE, purl = FALSE}
(cytokines <- read.csv('data/cytokines_3groups.csv'))
```

Check top (first 6 lines) using the function `head()`:

```{r, results='show', purl=FALSE}
head(cytokines)
```

The object `cytokines` is a `data.frame`.
A `data.frame` can be created by the functions `read.csv()` or `read.table()`.

### Data Frames

`data.frame` is the _de facto_ data structure for most tabular data and what we
use for statistics and plotting.

A `data.frame` is a collection of vectors of identical lengths. Each vector
represents a column, and each vector can be 
of a different data type (e.g. characters, integers, factors). The `str()`
function is useful to inspect the data types of the columns.

Now, generate final data.frame with Log2 transformed cytokine concentrations:

```{r, purl=FALSE}
cytokines_L <- data.frame(cytokines[,c(1:2)], log2(cytokines[,-c(1,2)]))
```

Examine the structure of this final `data.frame`:

```{r, purl=FALSE}
str(cytokines_L)  # reports the data types for each column
```

$ allow you to select columns using their names:

```{r, purl=FALSE}
cytokines_L$IL17F
cytokines_L$group
```

`cytokines_L$group` is a `factor` with 3 levels: "control", "infection_A", "infection_B"

### Factors

```{r, echo=FALSE, purl=TRUE}
### Factors
```

Factors are used to represent categorical data, and are an important class for statistical analysis and for plotting.

Factors contain a pre-defined set values, known as *levels*. 

By default, R always sorts *levels* in alphabetical order. It will assign `1` to the first level, `2` to the next, and so on. 

### Visualising and Analysing Cytokine Data

```{r, echo=FALSE, purl=TRUE}
### Visualising and Analysing Cytokine Data
```

To begin, let's `plot` our cytokine data:

```{r, purl=TRUE}
plot(cytokines_L[,-c(1,2)])
```

Add colors to distinguish the 3 groups:

```{r, purl=TRUE}
plot(cytokines_L[,-c(1,2)], 
     pch = 21, 
     bg = c("red", "green3", "blue")[unclass(cytokines_L$group)])
```

Focus in on 3 cytokines that look promising ("IL17F", "IL2", and "CXCL9"):
To begin, let's `plot` our cytokine data:

```{r, purl=TRUE}
plot(cytokines_L[,c("IL17F","IL2","CXCL9")], 
     pch = 21, 
     bg = c("red", "green3", "blue")[unclass(cytokines_L$group)])
```

Perform ANOVA:

```{r, purl=TRUE}
fit = aov(cytokines_L[,"IL17F"] ~ cytokines_L$group)

summary.aov(fit)
```

Box and Whisker Plot:

```{r, purl=TRUE}
boxplot(cytokines_L$IL17F~cytokines_L$group,
        ylab = "log.conc",
        xlab = "Groups",
        main = "IL17F serum levels in patients",
        col = "dark green")
```

Optional extra, create overlay:

```{r, purl=TRUE}
boxplot(cytokines_L$IL17F~cytokines_L$group,
        ylab = "log.conc",
        xlab = "Groups",
        main = "IL17F serum levels in patients",
        col = "dark green")
points(cytokines_L$IL17F~cytokines_L$group, 
         pch = 21,
         bg = "red")
```
